---
title: "R programming for data science-notes"
author: "Ai Zhong"
date: "8/15/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Five Basic or 'Atomic' Classes of Objects
* character
* numeric (real numbers)
* integer
* complex
* logical (True/False)

The most basic type of R object is a vector.  
To create an empty vector, use <span style="color:blue">*vector( )*</span> function, or use <span style="color:blue">*c( )*</span> function by concatenation.  
**A vector can only contain objects of the same class.**  
*Note: A list is represented as a vector, but can have objects of different classes*

### Numbers
If you want an integer, specify L suffix. eg: 1 is numeric, 1L is integer.

### Attributes
Attributes of an object can be accessed using <span style="color:blue">*attributes( )*</span> function.  
Not all R objects contain attributes.

### Explicit Coercion
Objects can be explicitly coerced from one class to another using the <span style="color:blue">*'as.\*'*</span> function.
*Note: to convert a datafram to matrix, use 'data.matrix( )'*.  
```{r}
x<-0:6
x
class(x)
as.character(x)
class(as.character(x))
```

### Matrices
Matrices are constructed *column-wise*. And matrices must have all element of the same class.  
```{r}
m<-matrix(1:6,nrow=2,ncol=3)
m
```
Matrices can also be created directly from vectors by adding a dimension attribute.  
```{r}
m<-1:10
dim(m)<-c(2,5)
m
```
Or Matrices can be created by row-binding <span style="color:blue">*rbind( )*</span> and column-binding <span style="color:blue">*cbind( )*</span>.  
```{r}
x<-1:3
y<-10:12
cbind(x,y)
rbind(x,y)
```
### List Creation
```{r}
x<-list(1,'a',TRUE,1+4i)
x

y<-vector('list',length=2)
y
```
### Factors
Factors: an integer vector where each integer has a *label*.  
Factors can be created with *factor( )* function.
```{r}
x<-factor(c('yes','yes','no','yes','no'))
x ## Levels are put in alphabetical order
table(x)
## to specify the level order:
y<-factor(c('yes','yes','no','yes','no'), levels=c('yes','no'))
y
table(y)
```
When you use *read.table( )*, usually factors is auto created for the dataset with characters or strings.  

## Missing Values
A *NaN* value is a type of *NA* but the converse is false.  
Use *is.na( )* and *is.nan( )* to test for missing values.

## Data Frames
Reading in datasets by *'read.table( )' and 'read.csv( )'* are usually dataframes. Or you can generate dataframe using *'data.frame( )'* function.  
```{r}
x<-data.frame(foo=1:4,bar=c(T,T,F,F))
x
```
### Names
**Summary:**  

Object     | Set column names | Set row names |
---------- | ---------------- |-------------- |
data frame | names( )         | row.names( )  |
matrix     | colnames( )      | rownames( )   |

## Subsetting R Object
### Subsetting a Vector

* '[' operator: return an object of the same class as the original. Can select multiple objects.  
* '[[' operator: extract elements of a list or a dataframe. Single element only, and may not be a list or dataframe.  
* '$' operator: extract elements of a list or dataframe by literal name. similar to '[['.  

Some Interesting Examples:  
```{r}
x<-c('a','b','c','c','d','a')
u<- x>'a'
u
x[u]
# or in a more compact way:
x[x>'a']
```
### Subsetting a Matrix
By default the extracted single element of matrix is returned as a vector of length 1, not a 1x1 matrix. This can be turned off by setting *'drop=FALSE'*.
```{r}
x<-matrix(1:6,2,3)
x[1,2]
x[1,2,drop=FALSE]
```
### Subsetting Lists

```{r}
x<-list(foo=1:4, bar=0.6)
x$foo # or x$bar
x[[1]] # or x[['foo']]
```
**Note: '[[' operator can be used with computed indices, '$' can only be used with literal names.**  
**Subsetting Nested Elements of a List**  
```{r}
x<-list(a=list(10,12,14), b=c(3.14,2.81), c=c('a','b'))
# get the 3rd element of the 1st element:
x[[c(1,3)]] 
x[[1]][[3]]
# use of '[' operator to extract multiple elements from a list:
x[c(1,3)] # get $a and $c from x
```
### Removing NA Values
Some Typical Examples:  
1.
```{r}
x<-c(1,2,NA,4,NA,5)
bad<-is.na(x)
x[!bad]
```
2.  
```{r}
x<-c(1,2,NA,4,NA,5,NA)
y<-c('a','b',NA,'d',NA,'f','e')
# note: x and y must be of the same length, and the value will follow x
# check limitations: ?complete.cases
good<-complete.cases(x,y)
good
x[good]
y[good]
```
## dplyr Package
Import the library  
```{r}
library(dplyr)
chicago<-readRDS('~/Desktop/StatswR/R Programming-JHU/chicago')
dim(chicago)
names(chicago)
```
####Select( )
**Use *'select( )'* to get the first 3 columns**
```{r}
subset<-select(chicago, city:dptp)
subset_neg<-select(chicago,-(city:dptp))
# equivalent code in base R would be:
# i<-match('city',names(chicago))
# j<-match('dptp',names(chicago))
# head(chicago[,-(i:j)],2)
head(subset,2)
head(subset_neg,2)
```
*Select( )* function also allows a special syntax that allows you to specify variable names based on patterns.  
Example:
```{r}
subset<-select(chicago,ends_with('2')) # or: starts_with('d')
str(subset)
```
#### Filter( )
Used to extract subsets of rows from a dataframe. Faster than *subset( )* in base R.  
```{r}
chic.f <- filter(chicago,pm25tmean2>30)
chic.f2 <- filter(chicago, pm25tmean2 > 30 & tmpd > 80)
str(chic.f)
```
```{r}
summary(chic.f$pm25tmean2)
```
#### Arrange( )
Used to reorder rows of a dataframe according to one of the variables/columns.  
Ascending order
```{r}
chicago<-arrange(chicago,date)
head(select(chicago, date, pm25tmean2),3)
tail(select(chicago, date, pm25tmean2),3)
```
Or descending order:
```{r}
chicago2<-arrange(chicago,desc(date))
head(select(chicago2, date, pm25tmean2), 3)
```
#### rename( )
Note: in base R, use names() to find column name first and then assign the desired new name to that element.  
```{r}
chicago<-rename(chicago,dewpoint=dptp,pm25=pm25tmean2)
head(chicago[,1:5],3)
```
#### mutate( )
Compute transformations of variables in a dataframe.  
Example: detrend pm25  
```{r}
chicago<-mutate(chicago,pm25detrend=pm25-mean(pm25,na.rm=TRUE))
head(chicago,2)
```
*transmute( )* function: same as mutate( ) but then drops all non-transformed variables.  
```{r}
head(transmute(chicago,
               pm10detrend=pm10tmean2-mean(pm10tmean2,na.rm=TRUE),
               o3detrend=o3tmean2-mean(o3tmean2,na.rm=TRUE)))
```

#### Group_by( )
Usually use *summarize( )* function after *group_by( )*.  
General operation is: split a dataframe into separate pieces defined by a variable/variables, then apply a summary function across those subsets.  
```{r}
# add a column 'year' to 'chicago' dataframe
chicago<-mutate(chicago,year=as.POSIXlt(date)$year+1900)
# create a separate dataframe that splits the original by year:
years<-group_by(chicago,year)
# compute summary statistics for each year in the dataframe:
summarize(years, pm25=mean(pm25,na.rm=TRUE),
          o3=max(o3tmean2,na.rm=TRUE),
          no2=median(no2tmean2,na.rm=TRUE))
```
#### %>% Pipeline operator
Useful for stringing together multiple dplyr functions in a sequence of operations.  
eg.: *third(second(first(x)))* is equal to *first(x)%>%second%>%third*  
Example: compute average pollutant level by month.(to see seasonal trends)  
```{r}
mutate(chicago, month=as.POSIXlt(date)$mon+1)%>%
        group_by(month)%>%
        summarize(pm25 = mean(pm25, na.rm=TRUE),
                  o3 = max(o3tmean2, na.rm=TRUE),
                  no2 = median(no2tmean2, na.rm=TRUE))
```

## Control Structures
### if-else
if(condition1){  
} else if(condition2) {  
} else {  
}

### for Loops  
Note: for one line loops, curly braces are not strictly necessary.  
**Nested for loops**  
Commonly for multidimensional or hierarchical data structures.   
Nesting beyong 2 to 3 levels often troublesome. Break up the loops using functions!  
**While Loops**  
Be careful it can be infinite!  
**Repeat Loops**  
Initiate an infinite loop from the start, use *break* to exit.  
Example: searching for a solution and do not stop until close enough.(so you don't know how many loops you need to go through at the beginning.)  

### Next, break  
*next* is used to skip an iteration of a loop.  
*break* is used to exit a loop immediately.  

## Split-Apply-Combine Strategy, *Apply Family

* lapply: loop over a list and evaluate a function on each element  
*Note: split( ) is often used in conjunction with lapply.*
* sapply: same as lapply but simplified result.   
* tapply: apply a function over subsets of a vector  
* mapply: a multivariate version of lappy  
* apply: apply a function over the margins of an array  
* vapply: is useful for checking of the output since you can specify the type of the output if the output is not as specified, error will occur (safer command than sapply)   

**Example 1 on lapply and sapply:**  
```{r}
# compare this two result:
lapply(1:3,function(x) x^2)
sapply(1:3,function(x) x^2)
# use of '...' argument:
x<-1:4
lapply(x,runif,min=0,max=10) #min=0 and max=10 are passed to runif()
```

**Example 2 on split( ) and sapply( ):**  
```{r}
library(datasets)
# use airquality dataset, split by month, there are 5 months in total(5-9)
s<-split(airquality,airquality$Month)
# Then we take column means for Ozone, solar.R, and Wind for each month
# Remember to remove NA values
sapply(s,function(x){colMeans(x[,c('Ozone','Solar.R','Wind')],
                              na.rm=TRUE)})

```

**Example 3 on tapply( ):**  
*similar to a combination of split() and sapply()*
```{r}
x<-c(rnorm(10),runif(10),rnorm(10,1))
f<-gl(3,10)
# sapply(split(x,f),mean) is equal to:
tapply(x,f,mean) # if you want a list, use tapply(x,f,mean,simplify=FALSE)
# when result is more than a single value, tapply will return a list automatically
tapply(x,f,range)
```

**Example 4 on apply():**  
Often used to rows/columns of a matrix or array.  
Using apply is not really faster than a loop, but it looks compact.
```{r}
x<-matrix(rnorm(200),20,10) # 20 rows and 10 columns
apply(x,2,mean) # take mean of each column
apply(x,1,mean) # take mean of each row
```
*Or one can use rowSums/rowMeans,colSums/colMeans as shortcuts.*

**Example 5 on mapply():**  
apply function over multiple R objects. 
function(FUN, ..., MoreArgs=NULL, SIMPLIFY=TRUE, USE.NAMES=TRUE)  
Note the argument order is different from others.  
```{r}
noise<-function(n,mean,sd){rnorm(n,mean,sd)}
mapply(noise, 1:5, 1:5, 2) # return 5 sets of random numbers, each with a different length and mean
```

The above mapply() is the same as:  
> list(noise(1, 1, 2),noise(2, 2, 2),noise(3, 3, 2),noise(4, 4, 2),noise(5, 5, 2))


## Simulation
### Generating Random Numbers  
For each probability distribution, there are 4 typical functions:  
* d for density
* r for random number generation
* p for cumulative distribution
* q for quantile function (inverse cumulative distribution)

### Setting the random number seed  
This ensures reporducibility of the sequence of random numbers.  
```{r}
set.seed(1)
rnorm(5)
```
If I call rnorm(5) again, I will get a different set of random numbers
```{r}
rnorm(5)
```
But if I call set.seed(1), it gives me back the previous numbers.  
```{r}
set.seed(1)
rnorm(5)
```

